---
title: "Tidying Up With ~~Marie Kondo~~ `purrr` et al."
subtitle: "<small>or: How I Learned to Stop Worrying and Love the Tidyverse</small>"
author: "Daniel Bride"

output: 
  revealjs::revealjs_presentation:
    highlight: haddock
    df_print: paged
    theme: solarized
    transition: slide
    self_contained: FALSE
    reveal_plugins: ["notes", "zoom"]
    slide_level: 1
    css: custom-reveal.css

---
```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(magick)
library(here)
source(here("src/create-fake-event-data.R"))
```

```{r imgs, include=FALSE}
usage <- 
  image_read("https://pbs.twimg.com/media/CfIflFzVAAAhWpQ.jpg") %>% 
  image_scale(geometry_size_percent(60))

garner_tweet <- image_read(here("resources/garner-tweet.PNG"))

map_diagram <- image_read("https://d33wubrfki0l68.cloudfront.net/f0494d020aa517ae7b1011cea4c4a9f21702df8b/2577b/diagrams/functionals/map.png") %>% 
  image_scale(geometry_size_percent(90))
  
map_df_diagram <- image_read("https://d33wubrfki0l68.cloudfront.net/12f6af8404d9723dff9cc665028a35f07759299d/d0d9a/diagrams/functionals/map-list.png")

map_types_table <- image_read(here("resources", "map-types-table.PNG"))
```

# Tidyverse Pt. II
- Part I covers many cool packages
  - https://github.com/slc-rug/tidyverse-explore
- I'll only cover one cool package: `purrr` 
- That's okay.

::: notes
- Part I covers many cool packages
- We're only covering one cool package today. That's okay.
- I'll use Tidyverse tools along the way.
:::

# Roadmap
- Tidyverse philosophy lesson
- Apply functions with `map()`, `modify()`
- Modify functions with `safely()`, `quietly()`, `possibly()`
- Reduce vectors with `reduce()`, `accumulate()`

::: notes
- At the risk of seeming uncool to advanced users, I'll simplify to a beginner level
- But first, an interesting diversion.
:::

# Interesting Diversion
<h2>An Impassioned Defense</h2>
`r garner_tweet`

::: notes
- Bryan Garner sticks up for what's right.
- Of all the people being outraged about all the things, he chose this.
- ¿Why does he care so much?
:::

# Interesting Diversion
<h2>Garner's other work</h2>
`r usage`

::: notes
- Garner's work is at the heart of a heated debate you probably didn't know existed: 
  - Descriptivism vs. Prescriptivism
:::

# Interesting(?) Diversion
<h2>Descriptive vs. Prescriptive Grammar</h2>
<div style="float: left; width: 50%;">
<h3>Descriptive:</h3>

- [Rules derived from usage]{.fragment .highlight-current-blue data-fragment-index="1"}
- [All usage equally valid]{.fragment .highlight-current-blue data-fragment-index="2"}
- [Comprehensive]{.fragment .highlight-current-blue data-fragment-index="3"}

</div>

<div style="float: right; width: 50%;">
<h3>Prescriptive:</h3>

- [Rules inform usage]{.fragment .highlight-current-blue data-fragment-index="1"}
- [One correct/best usage]{.fragment .highlight-current-blue data-fragment-index="2"}
- [Intentionally constrained]{.fragment .highlight-current-blue data-fragment-index="3"}

</div>

::: notes
- Stay with me.
:::

# Tidyverse Philosophy
<h2>Why Are We Talking About Grammar?</h2>
- Prescriptivism is [good for beginners]{.fragment .highlight-current-blue}
- Prescriptivism can [outsource cognitive burden]{.fragment .highlight-current-blue}
- Prescriptivism can [train a way of thinking]{.fragment .highlight-current-blue}
- [The Tidyverse is prescriptive]{.fragment .highlight-current-red}

::: notes
- **Good for beginners**: start with mimicry, diverge thoughtfully
- **Cognitive burden**: numerous alternatives sap brainpower
  - Ever dealt with a novel of a restaurant menu?
- Language drives thinking
:::

# Tidyverse Philosophy
<h2>Tidy Tendencies</h2>
- [Reuse existing data structures]{style="color: lightgray"}
- Compose simple functions with the pipe (`%>%`)
- Embrace functional (vs. imperative) programming
- Design for humans

[Reference: https://tidyverse.tidyverse.org/articles/manifesto.html]{style="float: right; font-size: small; padding: 50px 0;"}

::: notes
- Pipeable
- Functional
- Design
:::

# Learn to Love the Tidyverse
1. Try it
2. Ask why
3. Branch out confidently

::: notes
- Rules help master spoken language
- Rules help you master the R language
- Whether your first encounter with code
- Or you're bewildered at its quirky differences from other langs
:::

# The `purrr` Package (Finally!)
- A "functional programming toolkit" 
- Built for complex, iterative tasks
- `install.packages("tidyverse")`

::: notes
- Built to accomplish iterative tasks
- Designed to help you think through iterative tasks
:::

# Apply Functions
<h2>`map()`</h2>
- Do something to/with each element of a list or vector
- Return a list* of results
- <div class="fragment fade-up">\*Hold your horses. There are ways to return other types.</div>

# Apply Functions: `map()`
<h2>`map(`[`.x`]{.fragment .highlight-red data-fragment-index="1"}`, `[`.f`]{.fragment .highlight-blue data-fragment-index="2"}`, `[`...`]{.fragment .highlight-green data-fragment-index="3"}`)`</h2>
<h2>[Do something]{.fragment .highlight-blue data-fragment-index="2"} to/with each element of [a list or vector]{.fragment .highlight-red data-fragment-index="1"}</h2>
- [`.x`]{.fragment .highlight-red data-fragment-index="1"}: A vector or list
- [`.f`]{.fragment .highlight-blue data-fragment-index="2"}: 
  - A function name or formula to apply a function iteratively, or 
  - A vector or list to specify elements to extract from `.x`
- [`...`]{.fragment .highlight-green data-fragment-index="3"}: additional arguments passed to `.f`

# `map(.x, .f, ...)`
<h2>Visually:</h2>
`r map_diagram`

[Reference: https://adv-r.hadley.nz/functionals.html]{style="float: right; font-size: small; padding: 50px 0;"}

# `map(.x, .f, ...)`
<h2>Three Ways to Specify Functions</h2>
- `purrr`-style anonymous function shortcut (`~ mean(.x)`)
- Named function (`.f = mean`)
- Full function definition (`.f = function(x) mean(x)`)

# `map(.x, .f, ...)`
<h2>Let's make some unique example data</h2>
- Using [`babynames`](https://github.com/hadley/babynames) and [`lexicon`](https://github.com/trinker/lexicon)
- Unpopular baby names, popular sayings, inscrutable internet slang
```{r}
baby_pool <- 
  babynames::babynames %>% 
  group_by(sex, first_letter = str_sub(name, 1L, 1L)) %>% 
  top_n(-10, prop) %>% 
  pull(name) %>% 
  sort()

cliches <- lexicon::cliches

web_slang <- 
  lexicon::hash_internet_slang
```

::: notes
- Mention `set_names()` from `purrr`
- And `deframe()` from `tibble()`
:::

# `map(.x, .f, ...)`
<h2>Write function with `purrr` shortcut</h2>
<h3>Organize baby names by first letter</h3>
```{r}
chosen_babies <- 
  map(.x = LETTERS,
      .f = ~ {str_subset(baby_pool, paste0("^", .x))}) %>% 
  set_names(LETTERS)
```
- `~` is shorthand for function(x) { }
- When specifying this way, use `.` or `.x` to refer to data

::: notes
- Mention `set_names()` from `purrr`
:::

# `map(.x, .f, ...)`
<h2>Write named function</h2>
<h3>Output first three names for each letter</h3>

<small>- Notice we're using `...` to pass addtl argument (`3`) to `head()`</small>
```{r}
chosen_babies %>% map(head, 3)
```

# `map(.x, .f, ...)`
<h2>Write full function definition</h2>  
<h3>Match cliches to first letter</h3>
```{r}
alphabetical_cliches <- 
  LETTERS %>% 
  map(function(ltr) {
    cliches_sub <- str_subset(cliches, 
                              regex(paste0("^", ltr), ignore_case = TRUE))
    if(length(cliches_sub)) cliches_sub
    else "you're indescribable"}) %>% 
  set_names(LETTERS)
```

::: notes
- Probably a case where you'd want to define the function elsewhere, call it within map
:::

# ...Pardon the Interruption...
We'll Use This Later
```{r}
everything <- list(babies = chosen_babies, 
                   cliches = alphabetical_cliches, 
                   web_slang = deframe(web_slang)[1:26])
```

# `map(.x, .f, ...)`
<h2>Use `map()` to extract elements</h2>
- An integer n extracts nth item from each element of the list
```{r}
alphabetical_cliches %>% map(15, .default = "NOTHIN' TO SEE HERE!")
```

::: notes
- If you don't specify .default, you get an error if that index doesn't exist
:::
# `map(.x, .f, ...)`
<h2>Use `map()` to extract elements</h2>
- For named vectors/lists, a string extracts thus-named items from each
```{r}
everything %>% 
  map("Q", .default = "---No match---")
```

# `map(.x, .f, ...)`
<h2>Use `map()` to extract elements</h2>
- Use multiple values to index multiple levels
- Allowing you to access deeply nested values
- Combine characters and integers in a list
```{r}
everything %>% 
  map(list("Q", 3), .default = "***I got nothin'***")
```

# `map(.x, .f, ...)`
<h2>Returning other types</h2>
<h3 style="align: left">Obligatory `mtcars` example. Ugly version:</h3>

```{r}
map(mtcars, mean)
```

# `map(.x, .f, ...)`
<h2>Returning other types</h2>
- Just add a suffix: `map_*()` 
- E.g., `map_int()`, `map_lgl()`, ...
<h3>Better version:</h3>
```{r}
map_dbl(mtcars, mean)
```
- Names are conveniently preserved

# Apply Functions Across Two Lists/Vectors
<h2>`map2()`</h2>
- Do something with each element of two lists or vectors
- Return a list* of results
- <div class="fragment fade-up">\*Or character vector (`map2_chr()`), or double vector (`map2_dbl()`), or...</div>

# Apply Functions Across Two Lists/Vectors
<h2>`map(`[`.x`]{.fragment .highlight-red data-fragment-index="1"}`, `[`.y`]{.fragment .highlight-red data-fragment-index="1"}`, `[`.f`]{.fragment .highlight-blue data-fragment-index="2"}`, `[`...`]{.fragment .highlight-green data-fragment-index="3"}`)`</h2>
<h2>[Do something]{.fragment .highlight-blue data-fragment-index="2"} to/with each element of [two lists/vectors]{.fragment .highlight-red data-fragment-index="1"}</h2>
- [`.x`]{.fragment .highlight-red data-fragment-index="1"}: A vector or list
- [`.y`]{.fragment .highlight-red data-fragment-index="1"}: Another vector or list (length 1 or same as .x)
- [`.f`]{.fragment .highlight-blue data-fragment-index="2"}: A function name or formula to apply a function iteratively
- [`...`]{.fragment .highlight-green data-fragment-index="3"}: additional arguments passed to `.f`

::: notes
- The function needs to accept two arguments, or else why?
:::

# `map2(.x, .y, .f, ...)`
<h3>Celebrate the babies of the alphabet with alliterative cliches</h3>
```{r}
map2_chr(.x = chosen_babies, 
         .y = alphabetical_cliches,
         ~ paste0(str_to_sentence(sample(.y, 1, TRUE)), ", ",
                  str_to_sentence(sample(.x, 1)), "!"))
```

# So, `map3()`?
<h1>Of course not!</h1>

# Apply Functions With Many Arguments
<h2>`pmap()`</h2>
- Do something with each element of many lists/vectors
- Return a list or typed vector of results

# Apply Functions With Many Arguments
<h2>`pmap(`[`.l`]{.fragment .highlight-red data-fragment-index="1"}`,  `[`.f`]{.fragment .highlight-blue data-fragment-index="2"}`, `[`...`]{.fragment .highlight-green data-fragment-index="3"}`)`</h2>
<h2>[Do something]{.fragment .highlight-blue data-fragment-index="2"} to/with each element of [many lists/vectors]{.fragment .highlight-red data-fragment-index="1"}</h2>
- [`.l`]{.fragment .highlight-red data-fragment-index="1"}: A **list** of vectors/lists to supply as arguments to a function
  - All elements generally same length or length 1
- [`.f`]{.fragment .highlight-blue data-fragment-index="2"}: A function name or formula to apply a function iteratively
- [`...`]{.fragment .highlight-green data-fragment-index="3"}: additional arguments passed to `.f`

# `pmap(.l, .f, ...)`
```{r}
pmap(.l = everything, .f = function(babies, cliches, web_slang, ...) {
  paste0(str_to_sentence(web_slang), 
                         ", ",
                         str_to_sentence(sample(babies, 1, TRUE)), 
                         "--", 
                         sample(cliches, 1),
                         "!")
  })
```


# Using `map()` With Data Frames
`r map_df_diagram`

# Using `map()` With Data Frames
- (Fake) data similar to a study on cardiac events and air pollution
- For each event (case) in this set, need to compare PM2.5 (air pollution) on the same day of week in the same month
```{r}
ACS_events
```

# Using `map()` With Data Frames
<h2>List columns</h2>
```{r}
nested <- 
  ACS_events %>% 
  mutate(id_date = map(.x = ref_date,
                       .f = ~ {
                         row_refdate <- .x
                         month_dates <- seq(from = floor_date(.x, "month"),
                                            to = rollback(ceiling_date(.x, "month")),
                                            by = "day")
                         keep(month_dates, ~ wday(.x) == wday(row_refdate))
                       }))
```

# Using `map()` With Data Frames
<h2>List columns</h2>
- Each cell in `id_date` contains a list of 4 or 5 dates.
```{r, echo=FALSE}
nested
```

# Using `map()` With Data Frames
<h2>Nesting and unnesting</h2>
```{r}
unnested <- 
  nested %>% 
  unnest(id_date) %>% 
  arrange(id, id_date) %>% 
  mutate(event = ref_date == id_date)
unnested
```

# Using `map()` With Data Frames
<h2>(Here's how it ends):</h2>
```{r}
unnested %>% 
  left_join(PM25, by = c("id_date" = "date"))
```

# Modify a List
<h2>`modify(.x, .f, ...)`</h2>
- Behaves like `map()` 
- **Except** returns the type of object it receives

# Modify Functions
<h2>`safely()`, `quietly()`, `possibly()`, et al.</h2>
- "Adverbs": modify how functions ("verbs") behave
- Handy complement to `map()` and friends

# Modify Functions
- `safely(.f, otherwise = NULL, quiet = TRUE)` 
- Suppose I want to convert a bunch of things to dates
- If I happen to have some errant values, I want to know about them
- I don't want to stop everything because one date didn't work

# Modify Functions
- `safely(.f, otherwise = NULL, quiet = TRUE)` 
```{r}
bunch_o_ints <- list(1:2, "A", 4)

safe_date <- safely(.f = as_date, 
                    otherwise = "This position is empty")

map(bunch_o_ints, safe_date, origin = origin)
```

# Reduce/Accumulate
- `reduce(.x, .f, ..., .init .dir = c("forward", "backward"))`
- `accumulate(.x, .f, ..., .init, .dir = c("forward", "backward"))`
<h2>Simplest examples:</h2>
```{r}
reduce(1:10, `+`)

accumulate(1:10, `+`)
```

# Reduce/Accumulate
<h2>Bigger, more complex:</h2>
```{r, eval=FALSE}
list(df1, df2, df3, df4) %>% 
  reduce(.f = full_join)
```

# A Map of `map()` Varieties
`r map_types_table`

[Reference: https://adv-r.hadley.nz/functionals.html]{style="float: right; font-size: small; padding: 50px 0;"}

# Learn More
- https://r4ds.had.co.nz/iteration.html
- https://adv-r.hadley.nz/functionals.html
- https://purrr.tidyverse.org

